\mythesistitleNOTSYNPOSIS

### 2011-12-21 src/thesis.tex
Basic organization of doc. First real version of TOC. To be sent to n for comments.

\documentclass[MS]             % MS thesis dissertation.
              {iitmdiss}

@@

\def \mythesissubmissiondate {30 December 2011}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title page
\title{\mythesistitle} 
\author{\myname} 
\date{\MakeUppercase
  {\mythesissubmissionmonth}} 
\department{\MakeUppercase{\mydept}}

@@

\noindent 
This is to certify that the thesis titled {\bf \mythesistitle},
submitted by {\bf \myname}, to the Indian Institute of Technology,
Madras, for the award of the degree of {\bf \mydegree}, is
a bona fide record of the research work done by her under our
supervision.  The contents of this thesis, in full or in parts, have
not been submitted to any other Institute or University for the award
of any degree or diploma.

@@

\begin{abstract}
  \noindent {\bf Keywords}: \hspace*{0.5em} \parbox[t]{4.4in}{consecutive
    ones property, algorithmic graph theory, hypergraph isomorphism,
    interval labeling} \vspace*{24pt}

\def \tem {}
  \noindent 
  \annote{In this paper, we explore a natural generalization of
  results on binary matrices with the |\tem consecutive ones property|. 
  We consider the following constraint satisfaction problem. Given (i)
  a set system $\F \subseteq$ $(2^{U} \setminus \emptyset)$ of a
  finite set $U$ of cardinality $n$, (ii) a tree $T$ of size $n$ and
  (iii) a bijection called |\tem tree path labeling|, $\cl$ mapping the
  sets in $\cF$ to paths in $T$, does there exist at least one
  bijection $\phi:U \rightarrow V(T)$ such that for each $S \in \cF$,
  $\{\phi(x) \mid x \in S\} = \cl(S)$?  A tree path labeling of a set
  system is called |\tem feasible| if there exists such a bijection
  $\phi$.  We present an algorithmic characterization of feasible tree
  path labeling. COP is a special instance of tree path labeling
  problem when $T$ is a path.  We conclude with a polynomial time
  algorithm to find a feasible tree path labeling of a given set
  system when $T$ is a |\tem $k$-subdivided star|, set system has a
  single containment tree of overlap components and set size is
  limited to at most $k+2$. }{this is the abstract of the
  paper. expand to fit thesis.}
\end{abstract}

@@

\noindent 
\begin{tabbing}
xxxxxxxxxxx \= xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx \kill
\textbf{COP}   \> Consecutive Ones Property \\
\textbf{COT}   \> Consecutive Ones Testing \\
\textbf{ICPIA}   \> Intersection Cardinality Preservation Interval Assignment \\
\textbf{ICPPL}   \> Intersection Cardinality Preserved Path Labeling \\
% ADD MORE HERE
\end{tabbing}

@@

\section{Introduction}
\label{sec:intro}
\vspace{\secfirstparatrim}

Consecutive ones property is a non-trivial property of binary matrices
that has been studied widely in the literature for over past 50
years. Detection of COP in a matrix is possible efficiently and there
are several algorithms that achieve the same. This thesis documents
the work done on an extension of COP extended from the equivalent
interval assignment problem in \cite{nsnrs09}. These new results
rigorously prove a natural extension (to trees) of their
characterization as well as makes connections to graph isomorphism,
namely path graph isomorphism.

@@

\subsection{Illustration of the problem}
\label{sec:problem}
\vspace{\secfirstparatrim}

@@

%%%%%%%%%%%% THESIS UNCOMMENT %%%%%%%%%%%%
% \begin{figure}[t]%[htbp]
%  \centering
%   \includegraphics[scale=0.3]{../img/1_infinite_loop.png}
%   \caption{{\residenceblock} street map}
%   \label{fig:streetmap}
% \end{figure}

% \begin{figure}[t]%[htbp]
%   \centering
%   \includegraphics[scale=0.3]{../img/2_infinite_loop_BTWF.png}
%   \caption{{\residenceblock} street map with study group routes
%     allocated. Routes are color coded as follows: red for
%     \textcolor{red}{$\xLLL$} group, blue for \textcolor{blue}{$\xGGG$}
%     group, orange for \textcolor{YellowOrange}{$\xBBB$} group, green for
%     \textcolor{green}{$\xTTT$} group}
%   \label{fig:streetmappath}
% \end{figure}
% \begin{figure}[t] % [h]
%   \centering
%   \includegraphics[scale=0.5]{../img/3_infinite_loop.png}
%   \caption{Individual allocation of apartments to students in
%     {\residenceblock} that meets the requirements stated before.}
%   \label{fig:streetmappathpeople}
% \end{figure}
%%%%%%%%%%%% END THESIS UNCOMMENT %%%%%%%%%%%%
% \begin{figure}[htbp] %[t]%
%  \centering
%   (a) \includegraphics[scale=0.2]{../img/1_infinite_loop.png}
%   (b) \includegraphics[scale=0.2]{../img/2_infinite_loop_BTWF.png}
%  \label{fig:streetmap}
%   \caption{\figtabsize (a) {\residenceblock} street map (b) {\residenceblock} street map with study group routes
%     allocated. Routes are color coded as follows: red for
%     \textcolor{red}{$\xLLL$} group, blue for \textcolor{blue}{$\xGGG$}
%     group, orange for \textcolor{YellowOrange}{$\xBBB$} group, green for
%     \textcolor{green}{$\xTTT$} group}
% \end{figure}
\begin{figure}[htbp] %[t]%
  \centering
  \includegraphics[scale=0.3]{../img/3_infinite_loop.png}
  \caption{\figtabsize Individual allocation of apartments to students
    in {\residenceblock} that meets the requirements stated before.
    The routes are color coded as follows: red for
    \textcolor{red}{$\xLLL$} group, blue for \textcolor{blue}{$\xGGG$}
    group, orange for \textcolor{yellow}{$\xBBB$} group, green
    for \textcolor{green}{$\xTTT$} group. {\tiny {\em Peanuts images
        {\copyright} Charles Schulz}}}% YellowOrange
  \label{fig:streetmappathpeople}
\end{figure}
\tnote[TH]{make dashed/textured lines for routes. make it color agnostic.}

@@

The interval assignment problem to a set system is equivalent to the
consecutive ones property (COP) problem in binary matrices\cite{wlh02,
  nsnrs09}.  The COP problem is to rearrange rows (columns) of a
binary matrix in such a way that every column (row) has its {\un}s
occur consecutively. If this is possible the matrix is said to have
the COP.  COP is a well researched combinatorial problem and has
several positive results on tests for it and computing the COP
permutation (i.e. the course schedule in the above illustration) which
will be surveyed later in this document. Hence we are interested in
extensions of COP, more specifically, the extension of interval
assignment problem to tree path assignment problem (which is
illustrated by the study group accommodation problem).

Section~\ref{sec:background} gives a brief survey of COP and
optimization problems related to it followed by motivation for the
thesis in Section~\ref{sec:motive}.  Section~\ref{sec:results}
presents a summary of our results on the extension of COP namely, the
tree path labeling problem. \tnote[TH]{update for thesis}

\section{Consecutive Ones Property Testing - a Survey} %{Background}
\label{sec:background}
\vspace{\secfirstparatrim}
In this section, a brief survey of the consecutive ones problem and its
optimization problems is presented.

%%%%%%                              %%%%%%
%THX is the second level of ``tnote''-ing%
%%%%%%                              %%%%%%
\tnote[TH]{ADD: As it will be described in detail later in this
  document, isomorphism of certain % classes of graphs, namely chordal
%   graphs, have a close relationship with consecutive ones property and
%   generalizations of it.  This is perhaps because of how closely COP
%   of a matrix relates to properties of graphs derived from matrices as
%   seen in the following results.
}
\tnote[THX]{ADD: peo exists iff chordal. lexicographic BFS
  [tag:chordalGraph]} %
\tnote[THX]{ADD: A well known result in %perfect graph theory is that
%  the maximal cliques of  an interval graph $G$ can be linearly ordered
%   such that for all $v \in V(G)$, cliques containing $v$ are
%   consecutive in the ordering cite~{gh64}. This clearly means that a
%   graph $G$ is an interval graph if and only if
}% 
\tnote[THX]{verify
  from paper the statement of claim.}% 
\tnote[TH]{maximal clique vertex
  incidence matrix of%  $G$ has COP.  Also, maximal cliques of any
%   chordal graph can be enumerated in polytime $O(m+n)$
}
\tnote[THX]{citation?!!}%  
\tnote[TH]{cite~{fg65} uses these results to
  give the first polynomial time algorithm for COT.}%
\tnote[THX]{check. how do they use it?}%  
\tnote[TH]{ A bipartite graph
  is convex%  (on $R$) if and only if its half adjacency matrix has COP
%   on rows.  The results in cite~{bl76} on COT are based on the result
%   that interval graphs are AT-free chordal graphs.
}%
\tnote[THX]{the latter being Tucker's?}%  
\tnote[THX]{(2) TBD survey
  -- % see blue notes (in notebook) under Graph Isomorphism. namely
%   citations in cite:aas93 (3) a brief on heirarchy of chordal graphs,
%   path graphs, interval graphs, peo, clique tree etc. and results. (4)
%   interval graphs are incomparability graphs. see
%   Golumbic. [tag:classification] (5) finding min length hole in
%   bipartite graph is polynomial time [Sec 3.3.2 in cite:d08phd] (6)
%   theorem 2.2 in cite:d08phd - $G$ is union of v.d. caterpillars iff
%   $M$ has COP in rows. $M$ is edge vertex incidence matrix $(*,2)$ of
%   $G$
}%

\subsection{Matrices with COP}
\label{sec:copmatrices}
\vspace{\secfirstparatrim} 
As seen earlier, the interval assignment problem (illustrated as the
course scheduling problem in Section~\ref{sec:problem}), is a special
case of the problem we address in this thesis, namely the tree path
labeling problem (illustrated as the study group accommodation
problem). The interval assignment problem and COP problem are equivalent
problems. In this section we will see some of the results that
exists in the literature today towards solving the COP problem and
optimization problems surrounding it.

@@

The construction of PQ trees in \cite{bl76} draws on the
close relationship of matrices with COP to interval graphs. A PQ tree of a
matrix is one that stores all row (column) permutations of the matrix
that give the COP orders (there could be multiple orders of rows or columns) of the matrix. This
is constructed using an elaborate linear time procedure and is also
a test for planarity\tnote[GTC]{check check check. both interval
  graph and planarity in this paper?}. % Since this
% result, several improvements have been made to the procedural
% complexity of the algorithm itself and also derivative data structures
% that are easier to construct.
PQR trees is a generalized data structure based on PQ trees
\cite{mm96,mpt98}.  \cite{tm05} describes an improved algorithm to
build PQR trees. \tnote[GTC]{improv in terms of what?}\cite{wlh02}
describes the simpler algorithm for COT. Hsu also invented PC trees
\cite{wlh01}\tnote[TH]{This result first appeared inproc ISAAC92}
which is claimed to be much easier to implement. \cite{nsnrs09}
describes a characterization of consecutive ones property solely based
on the cardinality properties of the set representations of the
columns (rows); every column (row) is equivalent to a set that has the
row (column) indices of the rows (columns) that have one entries in
this column (row). This is interesting and relevant, especially to
this thesis because it simplifies COT to a great
degree. \tnote[GTC]{it reduces the solution search space. fill in the blanks.}

@@

\subsection{Optimization problems in COP}
\label{sec:optcop}
\vspace{\secfirstparatrim}  
So far we have been concerned about matrices that have the consecutive
ones property. However in real life applications, it is rare that data
sets represented by binary matrices have COP, primarily due to the
noisy nature of data available. At the same time, COP is not arbitrary
and is a desirable property in practical data representation
\cite{co98,jkckv04,k77}. In this context, there are several
interesting problems when a matrix does not have COP but is ``close''
to having COP or is allowed to be altered to have COP. These are the
optimization problems related to a matrix which does not have
COP. Some of the significant problems are surveyed in this section.

@@

\section{Generalization of COP - The Motivation}  %{Motivation}
\label{sec:motive}
\vspace{\secfirstparatrim}  
As seen in Section~\ref{sec:copmatrices}, \cite{nsnrs09} describes a
characterization of consecutive ones property based on the cardinality
properties of the set representations of the columns. This result is
very relevant to this thesis because aside from it simplifying COT to
a great degree, our generalization problem is motivated by their
results.

@@

\def \probdefwidth {0.85\linewidth}
\def \FTPL {{\sc Feasible Tree Path Labeling}}
\def \CFTPL {\sc Compute Feasible Path Labeling} 
\section{Summary of results}
\label{sec:results}
\vspace{\secfirstparatrim}
% We address the computation of TPL\footnote{Tree Path Labeling} for a
% given hypergraph\footnote{The term {\em hypergraph} is used instead
%   {\em set systems} are used interchangably in this
%   section. Hypergraphs that have TPLs on some target tree are called
%   {\em path hypergraphs}.} and target tree in this research. However
% optimization problems in TPL, akin to problems in
% Section~\ref{sec:optcop} for COP, is outside the scope of this thesis.
As we saw earlier by the result in \cite{nsnrs09}, pairwise
intersection cardinality preservation is necessary and sufficient for
an interval assignment to be feasible for a given
hypergraph\footnote{A {\em hypergraph} is an alternate representation
  of a set system and will be used through out this section.} and thus
is a characterization for COP. In our work we extend this
characterization and find that trio-wise intersection cardinality
preservation makes a tree path labeling \footnote{A {\em tree path
    labeling} $\cl$ is a bijection of paths from the target tree $T$
  to the hyperedges in given hypergraph $\cF$.} (TPL) feasible, which
is a generalization of the COP problem. This problem is defined as
follows.

@@

\def \kstar {$k$-subdivided star}
\def \CFTPLKTREE {\sc Compute $k$-subdivided Star Path Labeling}

@@

\section{Conclusion}
\label{sec:conclusion}
\vspace{\secfirstparatrim}

\tnote{ Some things to remember - have this somewhere.

  \begin{itemize}
  \item Examples: Illustration of the problem examples. both COP and
    extension
  \item Basic preliminaries: (definitions theorems etc) needed if any
    - graph theory
  \item Organization: Outline of document
  \end{itemize}

  Abstract must be a brief about what results we have and how it fits
  in the body of research.

  \begin{enumerate}
  \item \emph{Area:} - Broad to Specialized. i.e. Combinatorial
    algorithms -> Matrix reorganization -> general data reorganization
    (interval assignment) -> path assignment
  \item \emph{Class of problems:} say, data reorganization.
  \item \emph{Nature of results:} Is a generalization. We have a
    Polynomial algorithm for a subset of the generalization.
  \end{enumerate}

Introduction is a magnified version of Abstract with some brief
citations etc. mini survey. Here talk about canonization how it
connects these problems to graph isomorphism. Introduce our problem,
explain why it is important etc. How canonization is a
posisbility. Brief about results obtained.

Someone who doesn't want to go through the whole thesis must get the
gist of the whole document from the introduction chapter. that should
be the outlook.

once this is done, the rest of the chapters will fall into place.

{\em Survey chapter} is the full fledged expansion of the survey in
introduction with details, observations, theorems etc.

  {\em The main themes} of the thesis that must not be left out:
  \begin{enumerate}
  \item connection of COP to {\bf set systems} and thus to general
    data bases. \cite{nsnrs09}
  \item {\bf graph isomorphism} by canonization \cite{kklv10}
  \item {\bf certificate for a problem} \cite{mcc04}
  \end{enumerate}

Also see purple book notes.
}

\section{to area}
\section{to problem}
\section{to your thesis}
\section{Organization of document}

\chapter{Consecutive Ones property}
\tnote{survey of cop includes mainly 
\begin{enumerate}
\item \cite{d08phd} (perhaps minus chap 4?)
\item \cite{nsnrs09}
\item \cite{mcc04}
\item \cite{kklv10}
\end{enumerate}
.. to name a few.

sections will be as follows:}
\section{Characterization of COP}

\tnote{\cite{at72}, \cite{mcc04}, prime submatrix \cite{wlh02}?}
\section{Recognition of COP}
\subsection{\tnote{Polynomial time algorithms}}
\tnote{ (pq tree, pqr, icpia)}
\subsection{\tnote{Tucker's submatrices, Dom's algo to find them}}
\subsection{\tnote{McConnell's certificate}}

\section{Alteration to matrices to get COP}
\tnote{algorithms to make a matrix into one with COP
  (min/max-cos-r/c) \cite{d08phd} }
\section{brief analysis of hard problems with input having COP (chap
  5, 6 in dom)}
\label{sec:briefanal}
\tnote{Do I need this?}
\section{Complexity of certain COP variations}
\tnote{maybe combine with Section~\ref{sec:briefanal}\\
complexity - poly, logspace}

\chapter{Other problems related to COP }
\tnote{(see literature)}

\chapter{Research - Tree path labeling of path hypergraphs}
\tnote{Extension to trees. ( stuff in latin paper,)}

\tnote{UPDATE TO FINAL VERSION. BELOW IS OLDER VERSION WITH
  INCOMPLETE SEC 4.}
\section{Introduction}
Consecutive ones property (COP) of binary matrices is a widely studied
combinatorial problem. The problem is to rearrange rows (columns) of a
binary matrix in such a way that every column (row) has its $1$s occur
consecutively. If this is possible the matrix is said to have the COP.
This problem has several practical applications in diverse fields
including scheduling \cite{hl06}, information retrieval \cite{k77} and
computational biology \cite{abh98}.  Further, it is a tool in graph
theory \cite{mcg04} for interval graph recognition, characterization
of hamiltonian graphs, and in integer linear programming
\cite{ht02,hl06}.  Recognition of COP is polynomial time solvable by
several algorithms. PQ trees \cite{bl76}, variations of PQ trees
\cite{mm96,wlh01,wlh02,mcc04}, ICPIA \cite{nsnrs09} are the main ones.

%\noindent
The problem of COP testing can be easily seen as a simple constraint
satisfaction problem involving a system of sets from a universe. Every
column of the binary matrix can be converted into a set of integers
which are the indices of the rows with $1$s in that column. When
observed in this context, if the matrix has the COP, a reordering of
its rows will result in sets that have only consecutive integers. In
other words, the sets after reordering are intervals. Indeed the
problem now becomes finding interval assignments to the given set
system \cite{nsnrs09} with a single permutation of the universe (set
of row indices) which permutes each set to its interval. The result in
\cite{nsnrs09} characterizes interval assignments to the sets which
can be obtained from a single permutation of the rows.  They show that
for each set, the cardinality of the interval assigned to it must be
same as the cardinality of the set, and the intersection cardinality
of any two sets must be same as the interesction cardinality of the
corresponding intervals.  While this is naturally a necessary
condition, \cite{nsnrs09} shows this is indeed sufficient.  Such an
interval assignment is called an Intersection Cardinality Preserving
Interval Assignment (ICPIA).  Finally, the idea of decomposing a given
0-1 matrix into prime matrices to check for COP is adopted from
\cite{wlh02} to test if an ICPIA exists for a given set system.

\noindent 
{\bf Our Work.}  A natural generalization of the interval
assignment problem is feasible tree path labeling problem of a set
system. The problem is defined as follows -- given a set system $\cF$
from a universe $U$ and a tree $T$, does there exist a bijection from
$U$ to the vertices of $T$ such that each set in the system maps to a
path in $T$.  We refer to this as the {\em tree path labeling problem}
for an input set system, target tree pair -- $(\cF,T)$. As a special
case if the tree $T$ is a path, the problem becomes the interval
assignment problem.  We focus on the question of generalizing the
notion of an ICPIA \cite{nsnrs09} to characterize feasible path
assignments.  We show that for a given set system $\cF$, a tree $T$,
and an assignment of paths from $T$ to the sets, there is a feasible
bijection between $U$ and $V(T)$ if and only if all intersection
cardinalities among any three sets (not necessarily distinct) is same
as the intersection cardinality of the paths assigned to them and the
input runs a filtering algorithm (described in this paper)
successfully.  This characterization is proved constructively and it
gives a natural data structure that stores all the relevant feasible
bijections between $U$ and $V(T)$.  Further, the filtering algorithm
is also an efficient algorithm to test if a tree path labeling to the
set system is feasible.  This generalizes the result in
\cite{nsnrs09}.

>>>>

\documentclass[MS]             % MS thesis dissertation.
              {iitmdiss_as}    % Hacked by hyperbolicme.
%              {iitmdiss}      % If the powers to be are finicky.

@@

\def \mythesissubmissiondate {30 December 2011}
\def \mythesissubmissionmonth {\MakeUppercase {December 2011}}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title page
\title{\mythesistitleNOTSYNPOSIS} 
\author{\myname} 
\date{
  {\mythesissubmissionmonth}} 
\department{\MakeUppercase{\mythesisdept}}
%\department{\mythesisdept}

@@

\noindent 
This is to certify that the thesis titled {\bf \mythesistitleNOTSYNPOSIS},
submitted by {\bf \myname}, to the {\bf Indian Institute of Technology
Madras}, for the award of the degree of {\bf \mydegree}, is 
a bona fide record of the research work done by her under our
supervision.  The contents of this thesis, in full or in parts, have
not been submitted to any other Institute or University for the award
of any degree or diploma.

@@

\begin{abstract}
  \noindent {\em {\bf Keywords}: \hspace*{0.5em} \parbox[t]{4.4in}{consecutive
    ones property, algorithmic graph theory, hypergraph isomorphism,
    interval labeling}} \vspace*{24pt}

  \def \tem {}
  \noindent 
  Consecutive-ones property is a non-trivial property of binary
  matrices that has been studied widely in the literature for over
  past 50 years. Detection of COP in a matrix is possible efficiently
  and there are several algorithms that achieve the same. This thesis
  documents the work done on an extension of COP extended from the
  equivalent interval assignment problem in \cite{nsnrs09}. These new
  results rigorously prove a natural extension (to trees) of their
  characterization as well as makes connections to graph isomorphism,
  namely path graph isomorphism.

  \tnote{EXPAND. 
  Abstract must be a brief about what results we have and how it fits
  in the body of research.\\
  -- Area: Broad to Specialized. i.e. Combinatorial
    algorithms -> Matrix reorganization -> general data reorganization
    (interval assignment) -> path assignment\\
  -- Class of problems: say, data reorganization.\\
  -- Nature of results: Is a generalization. We have a
    Polynomial algorithm for a subset of the generalization.\\
% maybe add - we explore a natural
%     generalization of results on binary matrices with the |\tem
%     consecutive-ones property|.  We consider the following constraint
%     satisfaction problem. Given (i) a set system $\F \subseteq$
%     $(2^{U} \setminus \emptyset)$ of a finite set $U$ of cardinality
%     $n$, (ii) a tree $T$ of size $n$ and (iii) a bijection called
%     |\tem tree path labeling|, $\cl$ mapping the sets in $\cF$ to
%     paths in $T$, does there exist at least one bijection $\phi:U
%     \rightarrow V(T)$ such that for each $S \in \cF$, $\{\phi(x) \mid
%     x \in S\} = \cl(S)$?  A tree path labeling of a set system is
%     called |\tem feasible| if there exists such a bijection $\phi$.
%     We present an algorithmic characterization of feasible tree path
%     labeling. COP is a special instance of tree path labeling problem
%     when $T$ is a path.  We conclude with a polynomial time algorithm
%     to find a feasible tree path labeling of a given set system when
%     $T$ is a |\tem $k$-subdivided star|, set system has a single
%     containment tree of overlap components and set size is limited to
%     at most $k+2$.
  }
\end{abstract}

@@

\noindent 
\begin{tabbing}
xxxxxxxxxxx \= xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx \kill
\textbf{COP}   \> Consecutive-Ones Property \\
\textbf{COT}   \> Consecutive-Ones Testing \\
\textbf{ICPIA}   \> Intersection Cardinality Preservation Interval Assignment \\
\textbf{ICPPL}   \> Intersection Cardinality Preserved Path Labeling \\
% ADD MORE HERE
\end{tabbing}

@@

% \section{Introduction}
% \label{sec:intro}
% \vspace{\secfirstparatrim}

\tnote{ The main themes of the thesis that must not be left out:\\
  connection of COP to ||set systems|| and thus to general
  data bases. cite:nsnrs09\\
  ||graph isomorphism|| by canonization cite:kklv10\\
  ||certificate for a problem|| cite:mcc04\\
  Also see purple book notes.  }

Consecutive-ones property is a non-trivial property of binary matrices
that has been studied widely in the literature for over past 50
years. Detection of COP in a matrix is possible efficiently and there
are several algorithms that achieve the same. This thesis documents
the work done on an extension of COP extended from the equivalent
interval assignment problem in \cite{nsnrs09}. These new results
rigorously prove a natural extension (to trees) of their
characterization as well as makes connections to graph isomorphism,
namely path graph isomorphism.

\tnote{Have a para on Organization: Outline of document}
Section~\ref{sec:background} gives a brief survey of COP and
optimization problems related to it followed by motivation for the
thesis in Section~\ref{sec:motive}.  Section~\ref{sec:results}
presents a summary of our results on the extension of COP namely, the
tree path labeling problem. \tnote[TH]{update for thesis}

@@

\section{Illustration of the problem}
\label{sec:problem}
%\vspace{\secfirstparatrim}

@@

% \begin{figure}[t]%[htbp]
%  \centering
%   \includegraphics[scale=0.3]{../img/1_infinite_loop.png}
%   \caption{{\residenceblock} street map}
%   \label{fig:streetmap}
% \end{figure}

% \begin{figure}[t]%[htbp]
%   \centering
%   \includegraphics[scale=0.3]{../img/2_infinite_loop_BTWF.png}
%   \caption{{\residenceblock} street map with study group routes
%     allocated. Routes are color coded as follows: red for
%     \textcolor{red}{$\xLLL$} group, blue for \textcolor{blue}{$\xGGG$}
%     group, orange for \textcolor{yellow}{$\xBBB$} group, green for
%     \textcolor{green}{$\xTTT$} group}
%   \label{fig:streetmappath}
% \end{figure}
% \begin{figure}[t] % [h]
%   \centering
%   \includegraphics[scale=0.5]{../img/3_infinite_loop.png}
%   \caption{Individual allocation of apartments to students in
%     {\residenceblock} that meets the requirements stated before.}
%   \label{fig:streetmappathpeople}
% \end{figure}

\begin{figure}[htbp] %[t]%
 \centering
  (a) \includegraphics[scale=0.2]{../img/1_infinite_loop.png}
  (b) \includegraphics[scale=0.2]{../img/2_infinite_loop_BTWF.png}
 \label{fig:streetmap}
 \caption{\figtabsize (a) {\residenceblock} street map (b)
   {\residenceblock} street map with study group routes
   allocated. Routes are color coded as follows: red for
   \textcolor{red}{$\xLLL$} group, blue for \textcolor{blue}{$\xGGG$}
   group, orange for \textcolor{yellow}{$\xBBB$} group, green for
   \textcolor{green}{$\xTTT$} group}
% \end{figure}
% \begin{figure}[htbp] %[t]%
%   \centering
  \includegraphics[scale=0.5]{../img/3_infinite_loop.png}
  \caption{\figtabsize Individual allocation of apartments to students
    in {\residenceblock} that meets the requirements stated before.
    The routes are color coded as follows: red for
    \textcolor{red}{$\xLLL$} group, blue for \textcolor{blue}{$\xGGG$}
    group, orange for \textcolor{yellow}{$\xBBB$} group, green
    for \textcolor{green}{$\xTTT$} group. {\tiny {\em Peanuts images
        {\copyright} Charles Schulz}}}% YellowOrange
  \label{fig:streetmappathpeople}
\end{figure}
\tnote[TH]{make dashed/textured lines for routes. make it color agnostic.}

@@

The interval assignment problem to a set system is equivalent to the
consecutive-ones property (COP) problem in binary matrices\cite{wlh02,
  nsnrs09}.  The COP problem is to rearrange rows (columns) of a
binary matrix in such a way that every column (row) has its {\un}s
occur consecutively. If this is possible the matrix is said to have
the COP.  COP is a well researched combinatorial problem and has
several positive results on tests for it and computing the COP
permutation (i.e. the course schedule in the above illustration) which
will be surveyed later in this document. Hence we are interested in
extensions of COP, more specifically, the extension of interval
assignment problem to tree path assignment problem (which is
illustrated by the study group accommodation problem).

\section{Basic preliminaries - general definitions and nomenclature}
\tnote{(definitions theorems etc) needed if any
    - graph theory}

\section{Consecutive-Ones Property Testing - a Brief Survey} %{Background}
\label{sec:background}
%\vspace{\secfirstparatrim}
In this section, a brief survey of the consecutive-ones problem and its
optimization problems is presented.

%%%%%%                              %%%%%%
%THX is the second level of ``tnote''-ing%
%%%%%%                              %%%%%%
\tnote[TH]{ADD: As it will be described in detail later in this
  document, isomorphism of certain % classes of graphs, namely chordal
%   graphs, have a close relationship with consecutive-ones property and
%   generalizations of it.  This is perhaps because of how closely COP
%   of a matrix relates to properties of graphs derived from matrices as
%   seen in the following results.
}
\tnote[THX]{ADD: peo exists iff chordal. lexicographic BFS
  [tag:chordalGraph]} %
\tnote[THX]{ADD: A well known result in %perfect graph theory is that
%  the maximal cliques of  an interval graph $G$ can be linearly ordered
%   such that for all $v \in V(G)$, cliques containing $v$ are
%   consecutive in the ordering cite~{gh64}. This clearly means that a
%   graph $G$ is an interval graph if and only if
}% 
\tnote[THX]{verify
  from paper the statement of claim.}% 
\tnote[TH]{maximal clique vertex
  incidence matrix of%  $G$ has COP.  Also, maximal cliques of any
%   chordal graph can be enumerated in polytime $O(m+n)$
}
\tnote[THX]{citation?!!}%  
\tnote[TH]{cite~{fg65} uses these results to
  give the first polynomial time algorithm for COT.}%
\tnote[THX]{check. how do they use it?}%  
\tnote[TH]{ A bipartite graph
  is convex%  (on $R$) if and only if its half adjacency matrix has COP
%   on rows.  The results in cite~{bl76} on COT are based on the result
%   that interval graphs are AT-free chordal graphs.
}%
\tnote[THX]{the latter being Tucker's?}%  
\tnote[THX]{(2) TBD survey
  -- % see blue notes (in notebook) under Graph Isomorphism. namely
%   citations in cite:aas93 (3) a brief on heirarchy of chordal graphs,
%   path graphs, interval graphs, peo, clique tree etc. and results. (4)
%   interval graphs are incomparability graphs. see
%   Golumbic. [tag:classification] (5) finding min length hole in
%   bipartite graph is polynomial time [Sec 3.3.2 in cite:d08phd] (6)
%   theorem 2.2 in cite:d08phd - $G$ is union of v.d. caterpillars iff
%   $M$ has COP in rows. $M$ is edge vertex incidence matrix $(*,2)$ of
%   $G$
}%

\subsection{Matrices with COP}
\label{sec:copmatrices}
%\vspace{\secfirstparatrim} 
As seen earlier, the interval assignment problem (illustrated as the
course scheduling problem in Section~\ref{sec:problem}), is a special
case of the problem we address in this thesis, namely the tree path
labeling problem (illustrated as the study group accommodation
problem). The interval assignment problem and COP problem are equivalent
problems. In this section we will see some of the results that
exists in the literature today towards solving the COP problem and
optimization problems surrounding it.

@@

The construction of PQ trees in \cite{bl76} draws on the
close relationship of matrices with COP to interval graphs. A PQ tree of a
matrix is one that stores all row (column) permutations of the matrix
that give the COP orders (there could be multiple orders of rows or columns) of the matrix. This
is constructed using an elaborate linear time procedure and is also
a test for planarity\tnote[GTC]{check check check. both interval
  graph and planarity in this paper?}. % Since this
% result, several improvements have been made to the procedural
% complexity of the algorithm itself and also derivative data structures
% that are easier to construct.
PQR trees is a generalized data structure based on PQ trees
\cite{mm96,mpt98}.  \cite{tm05} describes an improved algorithm to
build PQR trees. \tnote[GTC]{improv in terms of what?}\cite{wlh02}
describes the simpler algorithm for COT. Hsu also invented PC trees
\cite{wlh01}\tnote[TH]{This result first appeared inproc ISAAC92}
which is claimed to be much easier to implement. \cite{nsnrs09}
describes a characterization of consecutive-ones property solely based
on the cardinality properties of the set representations of the
columns (rows); every column (row) is equivalent to a set that has the
row (column) indices of the rows (columns) that have one entries in
this column (row). This is interesting and relevant, especially to
this thesis because it simplifies COT to a great
degree. \tnote[GTC]{it reduces the solution search space. fill in the blanks.}

@@

\subsection{Optimization problems in COP}
\label{sec:optcop}
%\vspace{\secfirstparatrim}  
So far we have been concerned about matrices that have the consecutive
ones property. However in real life applications, it is rare that data
sets represented by binary matrices have COP, primarily due to the
noisy nature of data available. At the same time, COP is not arbitrary
and is a desirable property in practical data representation
\cite{co98,jkckv04,k77}. In this context, there are several
interesting problems when a matrix does not have COP but is ``close''
to having COP or is allowed to be altered to have COP. These are the
optimization problems related to a matrix which does not have
COP. Some of the significant problems are surveyed in this section.

@@

\section{Generalization of COP - The Motivation}  %{Motivation}
\label{sec:motive}
%\vspace{\secfirstparatrim}  
As seen in Section~\ref{sec:copmatrices}, \cite{nsnrs09} describes a
characterization of consecutive-ones property based on the cardinality
properties of the set representations of the columns. This result is
very relevant to this thesis because aside from it simplifying COT to
a great degree, our generalization problem is motivated by their
results.

@@

% \def \probdefwidth {0.85\linewidth}
% \def \FTPL {{\sc Feasible Tree Path Labeling}}
% \def \CFTPL {\sc Compute Feasible Path Labeling} 
\section{Summary of results}
\label{sec:results}
%\vspace{\secfirstparatrim}
% We address the computation of TPL\footnote{Tree Path Labeling} for a
% given hypergraph\footnote{The term {\em hypergraph} is used instead
%   {\em set systems} are used interchangably in this
%   section. Hypergraphs that have TPLs on some target tree are called
%   {\em path hypergraphs}.} and target tree in this research. However
% optimization problems in TPL, akin to problems in
% Section~\ref{sec:optcop} for COP, is outside the scope of this thesis.
As we saw earlier by the result in \cite{nsnrs09}, pairwise
intersection cardinality preservation is necessary and sufficient for
an interval assignment to be feasible for a given
hypergraph\footnote{A {\em hypergraph} is an alternate representation
  of a set system and will be used through out this section.} and thus
is a characterization for COP. In our work we extend this
characterization and find that trio-wise intersection cardinality
preservation makes a tree path labeling \footnote{A {\em tree path
    labeling} $\cl$ is a bijection of paths from the target tree $T$
  to the hyperedges in given hypergraph $\cF$.} (TPL) feasible, which
is a generalization of the COP problem. This problem is defined as
follows.

@@

% \def \kstar {$k$-subdivided star}
% \def \CFTPLKTREE {\sc Compute $k$-subdivided Star Path Labeling}

@@

\chapter{Consecutive-Ones property - Important Results from Literature}
\tnote{TBD: have a few lines about organization of chapter}

\tnote{ Survey chapter: is the full fledged expansion of the survey in
  introduction with details, observations, theorems etc.}

\section{Matrices with COP}
\tnote{Expand on ref:sec:copmatrices}

\section{Optimization problems in COP}
\tnote{Expand on ref:sec:optcop}

\chapter{Preliminaries to new results}
%\section{Preliminaries} 
\label{sec:prelims}

\tnote{TBD: have a few lines about organization of chapter}
